Lensing in the Schwarzschild Spacetime Applicable to Neutron Stars

### 2023-09-25 src/tex/ms.tex

For a neutron star that is observed to pulse with period $P$, observers at fixed
position on the surface follow coordinate trajectories that have 
\begin{eqnarray}
    \chi^\theta\left( \lambda \right) & = & \theta_0 \\
    \chi^r\left( \lambda \right) & = & \rNS \\
    \chi^\phi\left(\lambda \right) & = & \frac{2 \pi \chi^t\left(\lambda \right)}{P}
\end{eqnarray}
The last equation follows from the fact that $P$ is the proper time difference
between pulses arriving at infinity at fixed location on the celestial sphere;
and thus the \emph{coordinate} time difference between pulses is $\Delta t = P$.
(Beware!  This is not the \emph{proper} time between pulses as measured by
observers fixed to the neutron star surface; this latter quantity differs by a
gravitational redshift, a time dilation due to rotational motion, and a redshift
due to the radial acceleration of the rotational motion.)

@@

>>>>

For a neutron star that is observed to pulse with period $P$, observers at fixed
position on the surface follow coordinate trajectories in the body frame that
have 
\begin{eqnarray}
    \chi^\theta\left( \lambda \right) & = & \theta_0 \\
    \chi^r\left( \lambda \right) & = & \rNS \\
    \chi^\phi\left(\lambda \right) & = & \frac{2 \pi \chi^t\left(\lambda \right)}{P}
\end{eqnarray}
The last equation follows from the fact that $P$ is the proper time difference
between pulses arriving at infinity at fixed location on the celestial sphere;
and thus the \emph{coordinate} time difference between pulses is $\Delta t = P$.
(Beware!  This is not the \emph{proper} time between pulses as measured by
observers fixed to the neutron star surface; this latter quantity differs by a
gravitational redshift, a time dilation due to rotational motion, and a redshift
due to the radial acceleration of the rotational motion.)

@@

To convert this world-line tangent vector to the \emph{observer} coordinate
system, let the inclination of the neutron star spin be $\iota$ and the
obliquity (measered ccw from ``vertical'' in the observer frame) be $\omega$.
Then the neutron star spin three-vector has cartesian components in the observer
frame of 
\begin{equation}
    \vec{s} = \left( \cos \iota, -\sin \iota \sin \omega, \sin \iota \cos \omega \right).
\end{equation}
If the incoming ray strikes the surface at a point with colatitude $\theta_0$
and longitude $\phi_0$, then the vector pointing to the impact point has
cartesian components in the observer frame of 
\begin{equation}
    \vec{r}_0 = \left( \cos \phi_0 \sin \theta_0, \sin\phi_0 \sin \theta_0, \cos \theta_0 \right).
\end{equation}
The instantaneous world-line tangent of the observer fixed to the NS surface at
the point of impact is given in cartesian coordinates by 
\begin{equation}
    \vec{t}_0 = \frac{2 \pi r_{NS}}{P} \frac{\partial \chi^t}{\partial \lambda} \vec{s} \times \vec{r}_0. 
\end{equation}
In the observer spherical coordinate system the components of the tangent vector
are given by 
\begin{eqnarray}
    \frac{\mathrm{d}\chi^t}{\mathrm{d} \lambda} & \simeq & 1 + \frac{1}{2}  \vRot^2 \sin^2 \theta_0 + \frac{1 + \frac{3}{2} \vRot^2 \sin^2 \theta_0}{\rNS} \\
    \frac{\mathrm{d} \chi^r}{\mathrm{d} \lambda} & = & 0 \\
    \frac{\mathrm{d} \chi^\theta}{\mathrm{d} \lambda} & = & \frac{\mathrm{d} \chi^t}{\mathrm{d}\lambda} \frac{2\pi}{P} \left( - \cos\iota \sin \phi_0 - \cos\phi_0 \sin \iota \sin \omega \right) \\
    \frac{\mathrm{d} \chi^\phi}{\mathrm{d} \lambda} & = & \frac{\mathrm{d} \chi^t}{\mathrm{d}\lambda} \frac{2\pi}{P} \left( -\cos\iota \cos \phi_0 \cot \theta_0  + \sin \iota \left( \cos \omega + \cot \theta_0 \sin \phi_0 \sin \omega \right) \right)
\end{eqnarray}

[I have only worked out the geometry up to here.]
